---
title: "JHU Practical Machine Learning - Course Project"
author: "mb"
date: "July 26, 2015"
output: html_document
---

##BACKGROUND
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

##DATA
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.

##OBJECTIVES
The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. 

- create a report describing how you built your model
- how you used cross validation
- what you think the expected out of sample error is
- why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

## I. LOADING AND CLEANING THE DATA
Upon initial examination of the training data, based on the structure and summar there are several variables with missing values.
```{r, results='hide'}
#.csv files are stored in the working directory
training = read.csv("pml-training.csv")
str(training)
summary(training)
```

To address this, both the training and testing data will be loaded into R - adding a special string of characters that represent the missing entries using the na.strings option. 

All variables with missing values shall be omitted. 
```{r}
#loading data
training = read.csv("pml-training.csv", na.strings = c("NA","#DIV/0!",""))
testing = read.csv("pml-testing.csv", na.strings = c("NA","#DIV/0!",""))


#identifying variables without missing values
cols = colSums(is.na(training))
cols_nmv <- which(cols==0)

#create new data frames containing variables with complete entries
#removing irrelevant variables
training2 = training[,cols_nmv]
training2 = training2[,c(-5:-3,-1)]
testing2 = testing[,cols_nmv]
testing2 = testing2[,c(-5:-3,-1)]

#check dimensions of the data and take a look of the new dataset
dim(training2); dim(testing2)
head(training2); head(testing2)
```

## II. CROSS-VALIDATION
Using the Caret package, partitioning the training dataset for cross-validation.

Splitting the training dataset into 60% for subTrain and 40% for subTest.

SubTrain and SubTest datasets both contain 56 variables with 11776 and 7846 observations, respectively.

```{r}
library(caret)

set.seed(807)
inTrain = createDataPartition(y=training2$classe, p = 0.6, list = FALSE)
subTrain = training2[inTrain,]
subTest = training2[-inTrain,]


dim(subTrain); dim(subTest)
```


## III. PREDICTION MODEL
Building of prediction models.

### First algorithm used: 'rpart' or Decision tree. 
```{r}
set.seed(807)
#loading rpart packages
library(rpart)
library(rpart.plot)
library(caret)

#building the model using the subTrain data
mod1 = train(classe ~ ., method = "rpart", data = subTrain)
#results
mod1
mod1$finalModel
#simple plot
prp(mod1$finalModel)

#running predictions using the mod1 on subTest data  
pred1 = predict(mod1, subTest)

#confusion matrix for evaluation of the model
confusionMatrix(pred1, subTest$classe)

```
Accuracy is at 0.569. Not sufficient enough to give us confidence on our prediction model, thus we'll further explore using other algorith such as Random Forest.



### Second Algorithm used: 'rf' or Random Forest.

```{r, eval=FALSE}
set.seed(807)

#load random forest package
library(randomForest)

#predictors
predictors = names(cols_nmv[c(-5:-3,-1,-60)])
decision = names(cols_nmv[60])

#building model using the subTrain data
#mod2 = train(classe ~ ., method = "rf", data = subTrain)
mod2 = randomForest(subTrain[,predictors], subTrain[,decision])

#results
mod2
mod2$finalModel

#running predictions using mod2 on subTest data
pred2 = predict(mod2, subTest)

#model evaluation
confusionMatrix(pred2, subTest$classe)
```
The random forest model produced accuracy of 0.9976. 

## DECISION
Chose the random forest model as it performed better (with 0.9976 accuracy) than that of the decision tree model (0.569). The expected out-of-sample error is at 0.0024 or 0.24%.


Applying the random forest prediction model to the testing set.
```{r}
#applying the mod2 (random forest model) on our testing set 
pred3 = predict(mod2, testing)
pred3

```


## SUBMISSION
```{r}
#storing predictions to 'answers' variable
answers = pred3

#writing answers to text document
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}

pml_write_files(answers)

```

